Introduction_outline

### 2019-01-20 Write_up/Writing/Introduction_outline.Rmd
Updates

Globally, there are four major eastern boundary upwelling systems (EBUS). These are the Benguela Current off south-western Africa, the Canary Current off north-western Africa including its northern extension off the Iberian Peninsula of south-western Europe, the Peru-Humboldt Current off western South America, and the California Current off the western continental USA and north-western Mexico (Bakun et al, 2015; Pauly and Christensen, 1995). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans. EBUS cover only 1% of the world’s ocean surface, but despite these are among the most productive regions of the world. This is because these systems are subject to coastal upwelling, which is defined as the process whereby cold, nutrient rich, high CO2 concentrated, low pH, and low oxygen waters are pushed to the surface as a result of alongshore winds interacting with the earth’s rotation. As a result of upwelling, diverse marine fauna and flora are able to thrive in these areas because of the abundance of nutrients and favourable temperatures.

Upwelling is primarily caused by alongshore, equator ward winds. These winds are caused by cross-shore atmospheric pressure gradients, and these gradients occur predominantly during heating periods. Bakun then proposed that if there is an increase in greenhouse gases, it will cause more heating during the day and more cooling during the night which cause the temperature gradient to increase and would result in a strong pressure gradient, and as we know the pressure gradient will affect the wind and therefore will affect upwelling. Several people have tried to model this using different datasets of SST and wind pattern data at various resolution and initially found that increases in gases, such as CO2 causes a reduction in trade winds, and therefore a reduction in upwelling. However, these models were based on coarse data and when higher resolutions are used, such as for the Benguela EBUS, the models have shown that upwelling will intensify in accordance to the upwelling intensification hypotheses (eg, Hsieh and Boer 1992, Mote and Mantua 2002, Lima and Wethey’s 2012).

@@

This study focusses on the Benguela EBUS. The Benguela EBUS is unique in that it is bounded at both the equator ward and pole ward ends by warm water regimes. These currents are forced locally by the wind stress field off South-West Africa (Nelson and Hutchings 1983; Fennel 1997). This upwelling system stretches along the south-western coast of Africa from southern Angola to Cape Agulhas (Cole, 1999). Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S (Lutjeharms and Meeuwis, 1987). The main upwelling period in the Benguela north of 25°S extends mainly from March to November, peaking during August (Shannon, 1985), but upwelling persists throughout the year. Upwelling season in the northern Benguela extends between September and March, with very little upwelling taking place during winter (June-August). In the central Benguela and also at Cape Frio upwelling is continuous throughout the year (Shannon, 1985). The cold Benguela waters are bound in the south by the warm Agulus retroflexion region and in the north by the southward flowing Angola current. The flow of the Benguela is often considered to be topographically guided (Nelson and Hutchings, 1938; Barange et al. 1992). The Benguela system is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region (Shannon 1985; Cole 1999).

@@

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current (Peterson and Stramma, 1991). Research done by Yamagata and Iizuka (1995) discovered slight indication that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current (Duncombe, 1991). The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months (Boyd 1987; Shannon and Nelson 1996; Fennel 1997). Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation (Shannon 1985. Pitcher et al. 1992). Upwelling occurring along the South Coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind.  The wind increases somewhat away from the coast. This facilitated the idea of wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).

>>>>

Globally, there are four major eastern boundary upwelling systems (EBUS). These are the Benguela Current off south-western Africa, the Canary Current off north-western Africa including its northern extension off the Iberian Peninsula of south-western Europe, the Peru-Humboldt Current off western South America, and the California Current off the western continental USA and north-western Mexico (Bakun et al, 2015; Pauly and Christensen, 1995). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans. EBUS cover only 1% of the world’s ocean surface, but despite these are among the most productive regions of the world (Pauly and Christensen, 1995). This is because these systems are subject to coastal upwelling, which is defined as the process whereby cold, nutrient rich, high CO2 concentrated, low pH, and low oxygen waters are pushed to the surface as a result of alongshore winds interacting with the earth’s rotation (Huyer 1983). As a result of upwelling, diverse marine fauna and flora are able to thrive in these areas because of the abundance of nutrients and favourable temperatures.

Upwelling is primarily caused by alongshore, equator ward winds. These winds are caused by cross-shore atmospheric pressure gradients, and these gradients occur predominantly during heating periods. Bakun then proposed that if there is an increase in greenhouse gases, it will cause more heating during the day and more cooling during the night which cause the temperature gradient to increase and would result in a strong pressure gradient, and as we know the pressure gradient will affect the wind and therefore will affect upwelling. Several people have tried to model this using different datasets of SST and wind pattern data at various resolution and initially found that increases in gases, such as CO^2 causes a reduction in trade winds, and therefore a reduction in upwelling. However, these models were based on coarse data and when higher resolutions are used, such as for the Benguela EBUS, the models have shown that upwelling will intensify in accordance to the upwelling intensification hypotheses (eg, Hsieh and Boer 1992, Mote and Mantua 2002, Lima and Wethey’s 2012).

@@

This study focusses on the Benguela EBUS. The Benguela EBUS is unique in that it is bounded at both the equator ward and pole ward ends by warm water regimes. These currents are forced locally by the wind stress field off South-West Africa (Nelson and Hutchings 1983; Fennel 1999). This upwelling system stretches along the south-western coast of Africa from southern Angola to Cape Agulhas (Cole, 1999). Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S (Lutjeharms and Meeuwis, 1987). The main upwelling period in the Benguela north of 25°S extends mainly from March to November, peaking during August (Shannon, 1985), but upwelling persists throughout the year. Upwelling season in the northern Benguela extends between September and March, with very little upwelling taking place during winter (June-August). In the central Benguela and also at Cape Frio upwelling is continuous throughout the year (Shannon, 1985). The cold Benguela waters are bound in the south by the warm Agulus retroflexion region and in the north by the southward flowing Angola current. The flow of the Benguela is often considered to be topographically guided (Nelson and Hutchings, 1983; Barange et al. 1991a and 1991b). The Benguela system is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region (Shannon 1985; Cole 1999).

@@

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current (Peterson and Stramma, 1991). Research done by Yamagata and Iizuka (1995) discovered slight indication that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current (Duncombe, 1991). The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months (Boyd 1987; Shannon and Nelson 1996; Fennel 1999). Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation (Shannon 1985. Pitcher et al. 1992). Upwelling occurring along the South Coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind.  The wind increases somewhat away from the coast. This facilitated the idea of wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).
